---
title: "glimpse"
author: "Bayesian Outbreak Risk Assessment Tool"
output: html_document
---


```{r controls, include = FALSE}
require(ggplot2);require(dplyr);require(tidyr);library(cowplot);library(knitr);library(htmlTable)

## Set options
knitr::opts_chunk$set(echo = FALSE,warning = FALSE,message = FALSE)
options(scipen = 999, stringsAsFactors = FALSE)

## Pull in the data
load("fit.Rdata")
```

### Results

***
#### Methodological aspects
***

The analysis is based on a time-dependent susceptible-infectious-recovered (TSIR) model composed of two elements: (1) the distribution of the serial interval of the disease; and (2) a statistical model for the generation of observed secondary cases involving two main parameters: the basic reproduction number $R_0$ and the reporting rate $\rho$ (Riou et al, 2017). Fitting this model to weekly incidence data allows the estimation of these two parameters, which in turn will be used to stochastically simulate future incidence. From these simulations, several indicators of direct operational interest are then computed. The model and the simulations are conducted using *Stan* 2.15.1, a Bayesian statistical inference package with MCMC sampling (Carpenter et al, 2015).

***
#### Diagnostics
***

The sampling was based on __`r fit$nchains`__ chains with __`r fit$nit`__ iterations each,
including a burn-in of __`r fit$nwarmup`__ iterations, with a thinning ratio of __`r paste0("1:",fit$nthin)`__. A basic approach to detect the most common problems with the MCMC estimation process is to check (Gelman et al, 2014):

1. that the chains are well mixed after warm-up, which can be assessed visually by examining the *trace plot* of each parameter (Fig. 1A-C);

2. that autocorrelation (i.e. correlation between successive iterations) is low, which indicates a good efficiency of the sampling process. For lags >0, autocorrelation is expected increasingly low (Fig. 1D-F);

3. that the Gelman-Rubin convergence indicator $\hat{R}$ is close to 1 and lower than 1.05 (Fig. 1G and Table 1).

It is also always advisable to interpret the results with regards to the quality of the model fit with regards to the data (Fig. 1H). If no particular issue is detected in the diagnostics, the most common issues with MCMC sampling can generally be dismissed, and the results may be trusted. If any issue arises on the preceding three points, the results should not be trusted. In this case, increasing the number of iterations and the warm-up duration can resolve some situations. Increasing the thinning ratio can also be helpful in case of high autocorrelation.

<br>

<center> <b> <div style="max-width:500px;">
Figure 1. Post-estimation diagnostics: (A-C) trace plots of the three parameters of the model; (D-F) autocorrelation (AC) plots; (G) Gelman-Rubin Rhat; (H) model fit (mean and 50% and 95% confidence intervals).
</div> </b> </center>

```{r fig1, fig.height=7, fig.width=11,fig.align="center"}
f1 = plot_grid(fit$trace1 + scale_colour_discrete(guide=FALSE) + labs(x="Iterations") +
                 theme(axis.title=element_text(size=9),
                       axis.text=element_text(size=8)),
               fit$trace2 + scale_colour_discrete(guide=FALSE) + labs(x="Iterations") +
                 theme(axis.title=element_text(size=9),
                       axis.text=element_text(size=8)),
               fit$trace3 + scale_colour_discrete(guide=FALSE) + labs(x="Iterations") +
                 theme(axis.title=element_text(size=9),
                       axis.text=element_text(size=8)),
               ncol=3,
               labels=c("A","B","C"))

rhat = as.data.frame(fit$R_summarypars) %>%
  add_rownames() %>%
  ggplot(.) +
  geom_vline(aes(xintercept=c(1,1.05,1.1)),linetype=2,size=.5) +
  geom_segment(aes(x=1,xend=Rhat,y=rowname,yend=rowname)) +
  geom_point(aes(x=Rhat,y=rowname),size=2,shape=21,fill="skyblue") +
  scale_y_discrete(labels=c(expression(phi),expression(R[0]),expression(rho))) +
  scale_x_continuous(breaks=c(1,1.05,1.1)) +
  labs(x="Rhat",y="",title=NULL)


f2 = plot_grid(fit$acf1 +
                 theme_cowplot() +
                 labs(title=NULL,y=expression(paste("AC  (",R[0],")"))) +
                 theme(axis.title=element_text(size=9),
                       axis.text=element_text(size=8)),
               fit$acf2 +
                 theme_cowplot() +
                 labs(title=NULL,y=expression(paste("AC (",rho,")"))) +
                 theme(axis.title=element_text(size=9),
                       axis.text=element_text(size=8)),
               fit$acf3 +
                 theme_cowplot() +
                 labs(title=NULL,y=expression(paste("AC (",phi,")"))) +
                 theme(axis.title=element_text(size=9),
                       axis.text=element_text(size=8)),
               rhat +
                 theme_cowplot() +
                 theme(axis.title=element_text(size=9),
                       axis.text=element_text(size=8)),
               ncol=2,
               labels=c("D","E","F","G"))
f3 = plot_grid(f2,
               ggplot() +
                 geom_ribbon(data=fit$R_fit,aes(x=NWEEK,ymin=`25%`,ymax=`75%`),alpha=0.4,fill="skyblue") +
                 geom_ribbon(data=fit$R_fit,aes(x=NWEEK,ymin=`2.5%`,ymax=`97.5%`),alpha=0.3,fill="skyblue") +
                 geom_point(data=fit$data,aes(x=NWEEK,y=NCASES),shape=1,colour="black") +
                 geom_line(data=fit$R_fit,aes(x=NWEEK,y=mean),size=1) +
                 labs(x="Weeks",y="N") +
                 scale_y_continuous(expand=c(0,0),limits=c(0,max(fit$R_fit[["97.5%"]])*1.1)) +
                 theme_cowplot() +
                 theme(axis.title=element_text(size=9),
                       axis.text=element_text(size=8)),
               labels=c("","H"))

plot_grid(f1,f3,ncol=1,rel_heights = c(1,1.6))
```

***
#### Parameter estimates
***

The TSIR model includes two main parameters of interest: the reproduction number $R_0$ and the reporting rate $\rho$ (an additional parameter is $\phi$, the dispersion parameter). The parameters were given prior distributions, then, following Bayesian inference principles, these prior distributions were updated according using incidence data to obtain posterior distributions, which correspond to our current knowledge about the parameters. Eximining the differences between the prior and the posterior distributions allows to understand the respective influence of the prior distribution (that was chosen) and of the incidence data on the results (Fig. 2). Table 1 summarizes the posterior distributions of the parameters, together with the effective number of samples that were used in the estimation, and the Gelman-Rubin convergence indicator $\hat{R}$.

<br>

<center> <b> <div style="max-width:500px;">
Figure 2. Posterior distributions for $R_0$ (panel A) and $\\rho$ (panel B). The dashed lines represent the corresponding prior distributions.
</div> </b> </center>

```{r fig2, fig.height=4, fig.width=9,fig.align="center"}
dd = rbind(cbind(as.data.frame(do.call("cbind",fit$R_dens_r0[c("x","y")])),type="R0"),
           cbind(as.data.frame(do.call("cbind",fit$R_dens_rho[c("x","y")])),type="rho"))
ff = rbind(cbind(fit$prior_r0$ff,type="R0"),cbind(fit$prior_rho$ff,type="rho"))
f1 = ggplot() +
  geom_ribbon(data=filter(dd,type=="R0"),aes(x=x,ymax=y,ymin=0),alpha=0.7,fill="tomato") +
  geom_line(data=filter(dd,type=="R0"),aes(x=x,y=y),size=.7) +
  geom_line(data=filter(ff,type=="R0"),aes(x=x,y=prob),size=.7,linetype=2) +
  scale_y_continuous(expand=c(0,0)) +
  labs(title=NULL,x=expression(R[0]),y="Density") +
  theme_cowplot() +
  theme(axis.title=element_text(size=9),
        axis.text=element_text(size=8))
f2 = ggplot() +
  geom_ribbon(data=filter(dd,type=="rho"),aes(x=x,ymax=y,ymin=0),alpha=0.7,fill="tomato") +
  geom_line(data=filter(dd,type=="rho"),aes(x=x,y=y),size=.7) +
  geom_line(data=fit$prior_rho$ff,aes(x=x,y=prob),size=.7,linetype=2) +
  scale_y_continuous(expand=c(0,0)) +
  labs(title=NULL,x=expression(rho),y="Density") +
  theme_cowplot() +
  theme(axis.title=element_text(size=9),
        axis.text=element_text(size=8))
plot_grid(f1,f2,labels=c("A","B"))
```




<br>

<center> <b> <div style="max-width:500px;">
Table 1. Posterior distributions of the three parameters.
</div> </b> </center>

<center>

```{r tab1,results="asis"}
library(htmlTable)
nn = cbind(round(fit$R_summarypars[,c(1,3:8)],1),
                           round(fit$R_summarypars[,c(9)],0),
                           round(fit$R_summarypars[,c(10)],3))
dimnames(nn)[[1]] = c("$R_0$","$\\rho$","$\\phi$")
dimnames(nn)[[2]] = c("Mean","SD","2.5%","25%","50%","75%","97.5%","Eff. n","$\\hat{R}$")

htmlTable(nn)
```

</center>


***
#### Forward simulations
***

From the posterior distributions of $R_0$, $\rho$ and $\phi$, the remaining of the epidemic was simulated __`r fit$nchains*(fit$nit-fit$nwarmup)`__  times using a stochastic model mirroring the TSIR model. From the simulations we computed the predictive distribution of the weekly number of future incident cases to be reported by the surveillance system, summarized by a mean trajectory and a trajectory-wise 95% prediction band (Kolsrud, 2017).

<br>

<center> <b> <div style="max-width:500px;">
Figure 3. Predicted course of the epidemic from stochastic simulations (mean, 50% and 95% prediction intervals)
</div> </b> 


```{r fig4,fig.height=4, fig.width=6,fig.align="center"}
pathee = rbind(tail(fit$R_fit,1),fit$R_pred)
linee = filter(pathee,NWEEK==min(NWEEK))
ggplot() +
  geom_col(data=fit$data,aes(x=NWEEK,y=NCASES),fill=NA,colour="black",width=1) +

  geom_ribbon(data=pathee,aes(x=NWEEK,ymin=`25%`,ymax=`75%`),alpha=0.3,fill="tomato") +
  geom_ribbon(data=pathee,aes(x=NWEEK,ymin=`2.5%`,ymax=`97.5%`),alpha=0.3,fill="tomato") +
  geom_line(data=pathee,aes(x=NWEEK,y=mean),size=1,colour="tomato") +

  geom_vline(data=linee,aes(xintercept=NWEEK),size=0.3,linetype=2) +
  scale_y_continuous(expand=c(0,0) ) +
  labs(x="Weeks",y="N") +
  theme_cowplot() +
  theme(axis.title=element_text(size=9),
        axis.text=element_text(size=8))
```

</center>

***
#### Public health indicators
***

Parameter estimates and forecasts of the weekly number of clinical cases have limited value to public health decision-makers, as they do not directly translate to disease burden or strains on the healthcare system. Several indicators of direct operationnal value were thus computed from the simulated epidemic trajectories (Table 2):

1. the final epidemic size (i.e. the total number of clinical cases expected during the whole duration of the epidemic); 

2. the maximal of the future weekly incidence (for which the upper bound of the 97.5% prediction interval is of central interest);

3. the date of peak incidence, expressed as a number of weeks after the initial report; 

4. the date of the end of the period of high epidemic activity, also expressed as a number of weeks, and defined after __`r fit$w.eoo`__ consecutive weeks under a threshold of __`r fit$n.eoo`__ clinical cases.

<br>

<center> <b> <div style="max-width:500px;">
Table 2. Forecasts on four indicators with direct operational value.
</div> </b> 

```{r tab2,results="asis"}
ptotlp = format(round(fit$R_summarypred[1,c(1,4:8)]),big.mark=",")
maxinc = format(round(summarise(fit$R_pred,mean=max(mean),`2.5%`=max(`2.5%`),`25%`=max(`25%`),`50%`=max(`50%`),`75%`=max(`75%`),`97.5%`=max(`97.5%`))),big.mark=",")
peak = round(fit$R_peak[c(1,4:8)])
end = round(fit$R_end[c(1,4:8)])

mm = rbind(ptotlp,maxinc,peak,end)
dimnames(mm)[[1]] = c("Final epidemic size","Maximal weekly incidence","Week of peak incidence","Week of the end of high epidemic activity")
dimnames(mm)[[2]] = c("Mean","2.5%","25%","50%","75%","97.5%")

htmlTable(mm)
```

</center>

***
#### References
***

Carpenter B, Gelman A, Hoffman M, Lee D, Goodrich B, Betancourt M, et al. *Stan: a probabilistic programming language.* Journal of Statistical Software. 2015.

Gelman A, Carlin JB, Stern HS, Rubin DB. *Bayesian data analysis. vol. 2.* Chapman & Hall/CRC Boca Raton, FL, USA; 2014.

Kolsrud D. *Time-simultaneous prediction band for a time series.* Journal of Forecasting. 2007;26(3):171–188.

Riou J, Poletto C, Boëlle PY. *A comparative analysis of Chikungunya and Zika transmission.* Epidemics. 2017;19:43–52.

